The Design, Semantics, and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-10-07 paper.tex
Reorganizacao das figuras.
Revisao ate 3.3

In this work, we discuss the design of \CEU and propose a formal semantics for 
a small synchronous kernel that represents a subset of the language covering 
these new functionalities.
%
We also present an implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
Both use the \emph{ATmega328} with 32 Kbytes of FLASH and 2 Kbytes of SRAM:
\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
Considering the \VM back end, a simple application that blinks three LEDs 
periodically occupies less than 100 bytes and can be completely transmitted in 
4 radio messages, given that \emph{I/O} components are preloaded in the remote
devices.
In contrast, a \C-compiled version occupies more than 2 Kbytes, which includes 
all \emph{I/O} functionality~\cite{ceu.terra}.

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} discusses the design of \CEU, exposing its fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation back ends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

\CEU is a synchronous reactive language based on Esterel~\cite{esterel.ieee91} 
with support for multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events that are broadcast to all awaiting trails.
By synchronous, we mean that all trails at any given moment are either reacting 
to the current event or are awaiting another event;
in other words, trails are never reacting to different events.

@@

\begin{figure}
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

\CEU, following the Esterel mindset, has a strong imperative flavor, with 
explicit control flow through sequences, loops, parallels, and also 
assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be elegantly implemented.

@@

\begin{figure}
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence that the calling trail resumes and passes the 
assertion test (ln. 11).

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} is only valid for the next 
reaction, i.e., if an \code{await} and \code{emit} occur simultaneously in 
parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction inside loops (known as 
\emph{schizophrenic statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e; // w/o the restriction, awakes 1st trail and restarts the loop instantaneously
        await f;
    end
end
\end{lstlisting}

Even though both sides of the \code{par/or} have an \code{await} statement to 
avoid instantaneous termination, if the \code{await} could awake in the same 
reaction that reaches it, the \code{loop} would restart instantaneously 
resulting in infinite execution.

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute
        await e;  // await after
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
var bool should_execute = false;
par do
    loop do
        <...>
        if <...> then
            should_execute = true;
            emit e;
        end
    end
with
    loop do
        if should_execute then
            <...> // code to execute
        end
        await e;
    end
end
\end{lstlisting}
\end{minipage}
%
\caption{ Examples that circumvent the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop} (in the left), and by 
copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

Embedded applications make extensive use of shared memory, such as for sharing 
resources through memory-mapped registers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory.
%
Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{<<call 
f1() || call f2()>>}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
Therefore, Esterel forbids sharing memory between lines of execution: 
\emph{``if a variable is written by some thread, then it can neither be read 
nor be written by concurrent threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
reacts to a different input event (ln. 4,7) and, according to the semantics of 
\CEU, cannot occur simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
(ln. 4,7).

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (ln. 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 3);
the other periodically transmits the packet (ln. 5--18).
%
The periodic transmission is a loop that starts two other trails (ln. 6--17):
one handles the immediate retransmission request (ln. 7);
the other transmits the packet and waits for a confirmation (ln. 9--16).
%
%
The actual transmission (ln. 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping, in accordance with the specification.
Note that the call to \code{\_send\_enqueue}%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
(ln. 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
lexically-scoped packet (ln. 12).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion 
(ln. 15).
%
At any time, the client may request a retransmission (ln. 7), which terminates 
the \code{par/or} (ln. 6), aborts the ongoing transmission (ln. 14, if not 
idle), and restarts the loop (ln. 5).
%
The client may also request to stop the whole protocol at any time (ln. 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

@@

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program in the right of Figure~\ref{lst.timers}, although 
the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
it can at least ensure that the program always terminates with \code{v=1}:
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the main differences to Esterel, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and \C calls (which behave like in conventional 
imperative languages).

@@

Figure~\ref{lst.formal.syntax} shows the BNF-like syntax for a subset of \CEU 
that is sufficient to describe all semantic peculiarities of the language.
%
Except for $fin$ and all semantic-derived expressions (i.e., $awating$, 
$emitting$, and $p~@~loop; p$), which are discussed further, all other 
expressions map to their counterparts in the concrete language.

@@

We describe this relation with a set of small-step structural semantics rules, 
which are built in such a way that at most one transition is possible at any 
time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(await)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<=n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emit)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(pop)}
\end{align*}
}
%
An $await$ is simply transformed into an $awaiting$ (rule \textbf{await}) as an 
artifact to remember the external sequence number $n+1$ it can awake:
an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and it was reached in a previous 
reaction (i.e., sequence number $m<n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emit}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

@@

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   \\
\end{eqnarray*}
}
%
The deterministic behavior of the semantics relies on the \emph{isBlocked} 
predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
\textbf{and-adv2} and \textbf{or-adv2}.
These rules require the left branch $p$ to be blocked in order to allow the 
right transition from $q$ to $q'$.
%
Basically, the \emph{isBlocked} predicate determines that an expression becomes 
blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
expressions that cannot advance.

>>>>

In this work, we discuss the design of \CEU and propose a formal semantics for 
a small synchronous kernel that represents a subset of the language covering 
these new functionalities.
%
We also present an implementation of \CEU with two back ends:
one aiming for resource efficiency and interoperability with \C,
and another based on a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers.%
\footnote{
Arduino: \url{https://www.arduino.cc/en/Main/arduinoBoardUno}
\\
MICAz: \url{http://www.memsic.com/userfiles/files/Datasheets/WSN/micaz_datasheet-t.pdf}
\\
(Both use the \emph{ATmega328} microcontroller with 32 Kbytes of FLASH and 2 
Kbytes of SRAM.)}
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

In previous work~\cite{ceu.sensys13,ceu.terra}, we employed \CEU in the context 
of wireless sensor networks, developing a number of applications, protocols, 
and drivers.
%
We evaluated the expressiveness of \CEU in comparison to event-driven code in 
\C and attested a reduction in source code size (around 25\%) with a small 
increase in memory usage (around 5--10\% for \emph{text} and 
\emph{data})~\cite{ceu.sensys13}.
%
For the \VM back end, a simple application that blinks three LEDs periodically 
occupies less than 100 bytes and can be completely transmitted in 4 radio 
messages.
In contrast, the full image of a \C-compiled version occupies more than 2 
Kbytes~\cite{ceu.terra}.%
\footnote{
The \VM is preloaded in the devices with some \emph{I/O} functionality, whereas 
\C-based binaries are full images.
}

The rest of the paper is organized as follows:
Section~\ref{sec.ceu} discusses the design of \CEU, focusing on the fundamental 
differences to Esterel.
Section~\ref{sec.sem} presents a formal semantics for the control primitives of 
\CEU.
Section~\ref{sec.impl} presents the \C and \VM implementation back ends.
Section~\ref{sec.related} discusses other synchronous languages targeting 
embedded systems.
Section~\ref{sec.conclusion} concludes the paper.

@@

\CEU is a synchronous reactive language based on Esterel with support for 
multiple concurrent lines of execution known as \emph{trails}.
By reactive, we mean that programs are stimulated by the environment through 
input events that are broadcast to all awaiting trails.
By synchronous, we mean that all trails at any given moment are either reacting 
to the current event or are awaiting another event;
in other words, trails are never reacting to different events.

@@

\begin{figure}[b]
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
loop
   abort
      [
         await A
      $\|$
         await B
      ];
      emit O
   when R
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.48\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
// CEU
loop do
   par/or do
      par/and do
         await A;
      with
         await B;
      end
      emit O;
   with
      await R;
   end
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ A control specification implemented in Esterel and \CEU:
\emph{``Emit \code{O} after \code{A} and \code{B}, resetting each \code{R}.''}
\label{lst.abro}
}
\end{figure}

\CEU has a strong imperative flavor, with explicit control flow through 
sequences, loops, parallels, and also assignments.
Being designed for control-intensive applications, it provides support for 
concurrent lines of execution and broadcast communication through events.
%
In addition, \CEU also employs the synchronous model, in which programs advance 
in a sequence of discrete reactions to external events.
Internal computations within a reaction (e.g. expressions, assignments, and 
system calls) are considered to take no time in accordance with the synchronous 
hypothesis~\cite{rp.hypothesis}.
The \code{await} statements are the only ones that halt a running reaction and 
allow a program to advance in this notion of time.
%
To ensure that reactions run in bounded time and programs always progress, 
loops are statically required to contain at least one \code{await} statement in 
all possible paths~\cite{ceu.sensys13,esterel.primer}.
%
\CEU shares the same limitations with Esterel and synchronous languages in 
general:
computations that run in unbounded time (e.g., cryptography, image processing) 
do not fit the zero-delay hypothesis, and cannot be elegantly implemented.

@@

\begin{figure}[b]
\begin{minipage}[t]{0.43\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em,mathescape=true]
// ESTEREL
input A;    // external
signal B;   // internal
[[
    await A;
    emit B;
    call f();
$\|$
    await B;
    call g();
]]
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.53\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.1em]
// CEU
input void A;  // external (in uppercase)
event void b;  // internal (in lowercase)
par/and do
    await A;
    emit b;
    _f();
with
    await b;
    _g();
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Internal signals (events) in Esterel and \CEU: similar syntax, but 
different semantics.
\label{lst.prints}
}
\end{figure}

Internal events bring support for a limited form of subroutines, as depicted in 
Figure~\ref{lst.sub}.
The subroutine \code{inc} is defined as a loop (ln. 3--6) that continuously 
awaits its identifying event (ln. 4), incrementing the value passed as 
reference (ln. 5).
A trail in parallel (ln. 8--11) invokes the subroutine in reaction to event 
\code{A} through an \code{emit} (ln. 10).
Given the stacked execution for internal events, the calling trail pauses, the 
subroutine awakes (ln. 4), runs its body (yielding \code{v=2}), loops, and 
awaits the next ``call'' (ln. 4, again).
Only after this sequence the calling trail resumes and passes the assertion 
test (ln. 11).

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependency.
%
\CEU imposes a restriction that an \code{await} is only valid for the next 
reaction, i.e., if an \code{await} and \code{emit} occur simultaneously in 
parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous termination and 
re-execution of statements in the same reaction (known as \emph{schizophrenic 
statements}~\cite{esterel.schizo,esterel.schizo2}).
%
The example that follows relies on this restriction to avoid infinite 
execution:

\begin{lstlisting}[numbers=left,xleftmargin=2em]
event void e,f;
loop do
    par/or do
        await e;
    with
        emit e;     // w/o the restriction, the emit awakes 1st trail
        await f;    // and restarts the loop instantaneously
    end
end
\end{lstlisting}

Both sides of the \code{par/or} have an \code{await} statement to avoid 
instantaneous termination (ln. 4,7).
However, if the \code{emit} (ln. 6) could awake the \code{await} (ln. 4) in the 
same reaction that reaches them, the \code{par/or} would terminate and restart 
the \code{loop} instantaneously, resulting in infinite execution.

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute
        await e;  // await after
    end
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
var bool should_execute = false;
par do
    loop do
        <...>         // code that awaits
        if <...> then // some conditon
            should_execute = true;
            emit e;
        end
    end
with
    loop do
        if should_execute then
            <...>     // code to execute
        end
        await e;
    end
end
\end{lstlisting}
\end{minipage}
%
\caption{ Examples that circumvent the \emph{delayed await} restriction by 
post-fixing the \code{await} inside the \code{loop} (in the left), and by 
copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}

@@

Embedded applications make extensive use of global memory, such as for 
accessing resources through memory-mapped registers.
Hence, an important goal of \CEU is to ensure a reliable behavior for programs 
with concurrent lines of execution sharing memory.
%
Esterel is only deterministic with respect to reactive control: ``the same 
sequence of inputs always produces the same sequence of 
outputs''~\cite{esterel.primer}.
%
However, the execution order for operations with side-effects within a reaction 
is non-deterministic: ``if there is no control dependency, as in \code{<<call 
f1() || call f2()>>}, the order is unspecified and it would be an error to rely 
on it''~\cite{esterel.primer}.
%
Therefore, Esterel forbids sharing memory between lines of execution: 
\emph{``if a variable is written by some thread, then it can neither be read 
nor be written by concurrent threads''}~\cite{esterel.primer}.

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} (ln. 5,8) can never run concurrently, because each trail segment 
reacts to a different input event (ln. 4,7) and, according to the semantics of 
\CEU, cannot occur simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} (ln. 5,8) occur in the same reaction to input \code{A} 
(ln. 4,7).

\begin{figure}[b]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
the code in the left is safe because the trails access \code{x} atomically in 
different reactions;
the code in the right is unsafe because both trails access \code{y} in the same 
reaction.
\label{lst.shared}
}
\end{figure}

@@

To illustrate threats related to abortion, consider the unsafe example in the 
left of Figure~\ref{lst.fin}, which does not compile in \CEU.
It describes the state machine of a data collection protocol for sensor 
networks~\cite{wsn.ctp,ceu.sensys13}.
%
The input events \code{STOP}, \code{RETRANSMIT}, and \code{SENDACK} (ln. 1) 
represent the external interface of the protocol.
%
The protocol has to transmit a packet every minute, unless it receives a 
\code{RETRANSMIT} request to immediately re-transmit it, or a \code{STOP} 
request to terminate.
%
The protocol is composed of two main trails:
one simply monitors the stopping event (ln. 3);
the other periodically transmits the packet (ln. 5--18).
%
The periodic transmission is a loop that starts two other trails (ln. 6--17):
one handles the immediate retransmission request (ln. 7);
the other transmits the packet and waits for a confirmation (ln. 9--16).
%
%
The actual transmission (ln. 12--15) is enclosed with a \code{par/and} that 
takes at least one minute before looping (ln. 10), in accordance with the 
specification.
Note that the call to \code{\_send\_enqueue}%
\footnote{
The underline prefix marks (e.g., \code{\_send\_enqueue}) make interactions 
with external \C libraries explicit and are required in \CEU.
}
(ln. 14) is \emph{asynchronous}, handing to the radio driver a pointer to the 
lexically-scoped packet (ln. 12).
The driver makes the transmission in the background, holding the packet until 
it signals the application with the \code{SENDACK} to acknowledge completion 
(ln. 15).
%
At any time, the client may request a retransmission (ln. 7), which terminates 
the \code{par/or} (ln. 6), aborts the ongoing transmission (ln. 14, if not 
idle), and restarts the loop (ln. 5).
%
The client may also request to stop the whole protocol at any time (ln. 3).
%
Therefore, if the sending trail is aborted by the \code{STOP} or 
\code{RETRANSMIT} requests, the packet buffer goes out of scope, leaving behind 
a \emph{dangling pointer} in the radio driver, which will possibly transmit 
corrupted data.
%
%\subsubsection{Finalization}
%\label{sec.ceu.fin}

@@

\CEU also takes into account the fact that time is a physical quantity that can 
be added and compared.
For instance, for the program in the right of Figure~\ref{lst.timers}, although 
the scheduler cannot guarantee that the first trail terminates exactly in 11ms, 
it can at least ensure that the program always terminates with \code{v=1}.
%
Given that any non-awaiting sequence is considered to take no time in the 
synchronous model, the first trail is guaranteed to terminate before the second 
trail, because $10+1 < 12$.
%
A similar program in a language without first-class support for timers would 
depend on the execution timings for the code marked as \code{<...>}, making the 
reasoning about the execution behavior more difficult.

@@

In this section, we introduce a reduced syntax of \CEU and propose an 
operational semantics to formally describe the language.
We describe a small synchronous kernel with broadcast communication
highlighting the peculiarities of \CEU, in particular the stack-based
execution for internal events.
%
For the sake of simplicity, we focus on the control aspects of the language, 
leaving out side effects and \C calls (which behave like in conventional 
imperative languages).

@@

Figure~\ref{lst.formal.syntax} shows the BNF-like syntax for a subset of \CEU 
that is sufficient to describe all semantic peculiarities of the language.
%
Except for $fin$ and all semantic-derived expressions (i.e., $awating$, 
$emitting$, and $p~@~loop; p$), which are discussed further, all other 
expressions are equivalent to their counterparts in the concrete language.

@@

We describe this relation with a set of small-step structural semantics rules, 
which are built in such a way that at most one transition is possible at any 
time, resulting in deterministic reaction chains.
%
The transition rules for the primary expressions are as follows:
%
{ \setlength{\jot}{20pt}
\begin{align*}
\LL S,\1await(id) \RR &\ST
\LL S,\1awaiting(id,n+1) \RR
    & \textbf{(await)}      \\
%%%
\LL id:S,\1awaiting(id,m) \RR &\ST
\LL id:S,\1nop \RR, \2m<=n
    & \textbf{(awake)}   \\
%%%
\LL S,\1emit(id) \RR &\ST
\LL id:S,\1emitting(|S|) \RR
    & \textbf{(emit)}       \\
%%%
\LL S,\1emitting(|S|) \RR &\ST
\LL S,\1nop \RR
    & \textbf{(pop)}
\end{align*}
}
%
An $await$ is simply transformed into an $awaiting$ (rule \textbf{await}) as an 
artifice to remember the external sequence number $n+1$ it can awake:
an $awaiting$ can only transit to a $nop$ (rule \textbf{awake}) if its referred 
event $id$ matches the top of the stack and it was reached in a previous 
reaction (i.e., sequence number $m<=n$).
%
%Remember that in \CEU, the \code{await} statement returns the value associated 
%with the corresponding event: the yielded $mem$ represents the operation to 
%query that value.
%
An $emit$ transits to an $emitting$ holding the current stack level ($|S|$ 
stands for the stack length), and pushing the referred event on the stack (rule 
\textbf{emit}).
With the new stack level $|S|+1$, the $emitting(|S|)$ itself cannot transit, as 
rule \textbf{pop} expects its parameter to match the current stack level.
This trick provides the desired stack-based semantics for internal events.

@@

Proceeding to parallel compositions, the semantic rules for $and$ and $or$ 
always force transitions on their left branches $p$ to occur before their right 
branches $q$:
%
{ \setlength{\jot}{20pt}
\begin{eqnarray*}
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p'~and~q) \RR }
    & \textbf{(and-adv1)}      \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~and~q) \RR \ST \LL S', (p~and~q') \RR }
    & \textbf{(and-adv2)}      \\
%%%
& \frac
    {\DS \LL S,p \RR \ST \LL S',p' \RR }
%   -----------------------------------------------------------
    {\DS \LL S, (p~or~q) \RR \ST \LL S', (p'~or~q) \RR }
    & \textbf{(or-adv1)}   \\
%%%
& \frac
    {\DS isBlocked(n,S,p) \1,\2 \LL S,q \RR \ST \LL S',q' \RR }
%   -----------------------------------------------------------
    {\DS \LL S (p~or~q) \RR \ST \LL S', (p~or~q') \RR }
    & \textbf{(or-adv2)}   %\\
\end{eqnarray*}
}
%
The deterministic behavior of the semantics relies on the \emph{isBlocked} 
predicate, which is defined in Figure~\ref{fig.isBlocked} and used in rules 
\textbf{and-adv2} and \textbf{or-adv2}.
These rules require the left branch $p$ to be blocked in order to allow the 
right transition from $q$ to $q'$.
%
Basically, the \emph{isBlocked} predicate determines that an expression becomes 
blocked when all of its trails in parallel hang in $awaiting$ and $emitting$ 
expressions that cannot advance.

\begin{figure}
{\small
\begin{align*}
  isBlocked(n,a:S, awaiting(b,m)) &= (a \neq b \1\vee\1 m > n)   \\
  isBlocked(n,S, emitting(s))    &= (|S| \neq s)                     \\
  isBlocked(n,S, (p~;~q))        &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~@~loop~q))   &= isBlocked(n,S,p)             \\
  isBlocked(n,S, (p~and~q))      &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, (p~or~q))       &= isBlocked(n,S,p) \wedge
                                    isBlocked(n,S,q)             \\
  isBlocked(n,S, \_)             &= false \2  (mem,await,      \\
                                  &    \5\5\5\2 emit,break,if,loop)   %\\
\end{align*}
}%
%\rule{14cm}{0.37pt}
\caption{
The recursive predicate $isBlocked$ is true only if all branches in parallel 
are hanged in $awaiting$ or $emitting$ expressions that cannot transit.
\label{fig.isBlocked}
}
\end{figure}

%%%
